---
title: "Red Spruce Selection"
author: "Nicole Phelan"
date: "2025-09-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir="~/projects/eco_genomics_2025/population_genomics/myresults/ANGSD/PCA_ADMIX/") 
```

Loading in the libraries in R

```{r}
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(qqman)
```

Re-making the scree plot for just red spruce samples after removing the black spruce

```{r}
allRScov <- as.matrix(read.table("allRS_poly_K3.cov"))

allRSPCA <- eigen(allRScov)

var <- round(allRSPCA$values/sum(allRSPCA$values),3)

var[1:3]

# A "screeplot" of the eigenvalues of the PCA:

barplot(var, 
        xlab="Eigenvalues of the PCA", 
        ylab="Proportion of variance explained")
```

Prepare the sample ID information for plotting later

```{r}
names <- read.table("allRS_bam.list")
names <- unlist(strsplit(basename(as.character(names[,1])), split = ".sorted.rmdup.bam"))
split = strsplit(names, "_")
pops <- data.frame(names[1:95], do.call(rbind, split[1:95]))
names(pops) = c("Ind", "Pop", "Row", "Col")
```

Make the PCA on just the red spruce samples

```{r}
data = as.data.frame(allRSPCA$vectors)
data = data[,c(1:3)]
data = cbind(data, pops)

# PCA with just red spruce (95 samples):

ggscatter(data, x = "V1", y = "V2",
          color = "Pop",
          mean.point = TRUE,
          star.plot = TRUE,
          main="Red spruce-Black spruce genetic PCA") +
  theme_bw(base_size = 13, base_family = "Times") +
  labs(x = paste0("PC1: (",var[1]*100,"%)"), y = paste0("PC2: (",var[2]*100,"%)"))
```

How do our PC1 scores compare to the ancestry scores from the Admixture analysis? 

Test for correlation between Admixture scores from the red-black spruce analysis and the PC1 scores from the red spruce only analysis. Is PC1 still and axis of introgression? Is there any evidence for introgression along PC2.

```{r}
q_K2 <- read.table("RSBS_poly_K2.admix.2.Q")
names(q_K2) = c("qBS","qRS")

data2 <- cbind(q_K2[1:95,1:2],data)
  
p1 <- ggscatter(data2, x ="qBS", y = "V1",
          color="Pop",
          main="Red spruce-Black spruce genetic PCA") +
  theme_bw(base_size = 13, base_family = "Times") +
  labs(x = "Black spruce ancestry", y = "PC1 score")

p2 <- ggscatter(data2, x ="qBS", y = "V2",
                color="Pop",
                main="Red spruce-Black spruce genetic PCA") +
  theme_bw(base_size = 13, base_family = "Times") +
  labs(x = "Black spruce ancestry", y = "PC2 score")

ggarrange(p1,p2)
```

Read in selection statistics and calculate p-values based on the chi square distribution.

```{r}
s <- read.table("allRS_poly_K3.selection")

# convert test statistic to p-value
pval_PC1 <- as.data.frame(1-pchisq(s[,1],1))
names(pval_PC1) = "p_PC1"

pval_PC2 <- as.data.frame(1-pchisq(s[,2],1))
names(pval_PC2) = "p_PC2"

## read positions
# pos <- read.table("allRS_poly_K3.sites",sep="\t",header=T, stringsAsFactors=T)
# dim(pos)
```

Read the data on the SNP positions themselves

```{r}
## read in site positions

PCAngsd_sites <- read.table("allRS_poly_K3.sites")
MAF_all_sites <- read.table("/gpfs1/cl/ecogen/pbio6800/PopulationGenomics/ANGSD/allRS_poly.mafs.gz", header=T)

pos_filtered <- cbind(MAF_all_sites,PCAngsd_sites)[which(PCAngsd_sites==1),]

dim(pos_filtered)
```

Generating a Manhattan plot

```{r}
plot(-log10(pval_PC1$p_PC1), col="black", cex=0.8, xlab="Locus position", main="Selection outliers: PCAngsd PC1")

plot(-log10(pval_PC2$p_PC2), col="maroon", cex=0.8, xlab="Locus position", main="Selection outliers: PCAngsd PC2")
```

